# Chapter 6: Analysis of longitudinal data

## 1. Analyses of Chapter 8 of MABS using the RATS data
Reading in the data, setting the Group and ID variables to be factors and checking the data.

The variables are as follows:

ID - the ID number of each rat

Group - which of three different diets each rat was on

WD - day of measurement starting with the first day as 1

Weight - the body weight of the rat in grams

Time - as WD above but as an integer (with WD removed from the start)

```{r}
RATSL <- read.csv("data/RATSL.csv", sep = ",")
RATSL$Group <- factor(RATSL$Group)
RATSL$ID <- factor(RATSL$ID)

names(RATSL)

# Look at the structure of rats
str(RATSL)

# Print out summaries of the variables
summary(RATSL)
```


The following methods are for inital exploration and inital analysis of this longitudinal dataset.

## Graphs or analysis results 
These graphs can provide insights into the interesting patters of response over time and the structure of treatment differences. They can also help identify possible outliying observaitons. 
We're tryign to expose patterns in the data with the following guidlines in mind:
- show as much raw data as possible
- highlight aggregate pattersns
- identify both corss-sectional and longitudinal patterns
- try to make the idenification fon unsual individuals or unusual observations simple

### Preliminary assessment:
Below is a plot of the weight values for all rats over time differentiating between the three groups by line type (unbroken, dotted and dashed). Some observations from the plot: 

 - The generally increases over the period of the study.  

 - The rats in groups 2 and 3 have much heavier weights than those in group 1 and also increase in weight over the period of the study by much more than those in group 1.

```{r, include=FALSE}
library(dplyr, tidyr)
library(ggplot2)
```

```{r, fig.width=10, fig.height=10, warning=FALSE}

# Draw the plot of measures of weight over time
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) + scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) + scale_y_continuous(name = "Weight (grams)") + theme(legend.position = "top")

```

### Standardized values
To see a plot that reduces the impact of the tracking phenomenon we can make a plot of standardized values. The standardized values are obtained by subtracting the mean from the original observation then dividing by the corresponding weight measure standard deviation.

```{r, fig.width=10, fig.height=10, warning=FALSE}
# Standardise the variable bprs using the above equation
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()
 
# Glimpse the data
glimpse(RATSL)
 
# Plot again with the standardised bprs
ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rat weights")
```

### Summary Graphs
Showing average (mean) profiles for each diet group along with some indication of the variation of the observations at each time point, in this case the standard error of mean (showing the precision of the estimate of the mean).

We can see in the plot below that there is a considerable gap with no overlap between the mean profiles of the three groups suggesting significant difference between the groups with respect to the mean Weight values.

```{r,  warning=FALSE}
# Number of weeks, baseline (week 0) included
n <- RATSL$Time %>% unique() %>% length()
 
# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()
 
# Glimpse the data
glimpse(RATSS)
 
# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group, color = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = "right") +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

### Side by Side box plots and summary measure
Below is an additional comparison using side-by-side box plots with the repeated measurements of weight transformed into one value, the mean, for each rat. The boxplots show the greater variabilitiy of the second and third treatment groups as well as a bit of skew. One point of group 2 seems like an outlier being far above the majority of values, a second boxplot follows with that outlier removed. It doesn't seem like the outlier was biasing the conclusions. Even without it the groups maintain their difference and relative positions.

```{r, warning=FALSE}
# Create a summary data by treatment and subject with mean as the summary variable
RATSLSum <- RATSL %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
 
# Glimpse the data
glimpse(RATSLSum)
summary(RATSLSum)
 
# Draw a boxplot of the mean versus treatment
ggplot(RATSLSum, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), over time")
 
# Create a new data by filtering the outlier
RATSLSum2 <- filter(RATSLSum, RATSLSum$mean < 590)
 
 
# Glimpse the data
glimpse(RATSLSum2)
 
# Draw a boxplot of the mean versus treatment
ggplot(RATSLSum2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), over time")

```


### A formal test for difference
Chapter 8 of MABS included a t-test but our dataset has three groups.

The Anova table below using a baseline shows us no significant difference between the groups so it would seem that the difference between rat weights is mostly due to their starting weights rather than the diet having an impact. This is not something we could tell from the previous boxplots of the summary measure analysis.

```{r}
RATS <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep = "\t")
 
# Adding the baseline from the original data as a new variable to the summary data
RATSLSum3 <- RATSLSum %>%
  mutate(baseline = RATS$WD1)
 
# Fitting the linear model with the mean as the response/target and baseline + treatment as the response from the RATSL8S1 
fit <- lm(mean ~ baseline + Group, data = RATSLSum3)
 
# Compute the analysis of variance table for the fitted model with anova() and pay close attention to the significance of baseline
anova(fit)

```


### Missing values
Chapter 8 of MABS for IODS includes a section on dealign with missing values but the following check indicates that our data has no missing values.

```{r}
summary(complete.cases(RATSL))
```




## 2. Analyses of Chapter 9 of MABS using the BPRS data.
A group of 40 men where randomly put into two different groups that were given different treatments and tested over 8 weeks (week 0 was before treatment began) for their Brief Psychiatric Rating Scale (BPRS) score which is used to evaluate patients suspected of having schizophrenia.

We're looking to determine the explanatory variables most associated with the change in BPRS. The below analysis is more complete and more appropriate for longitudinal data than section 1 above. 

### Graphs or analysis results
### Interpretations

```{r}
BPRSL <- read.csv("data/BPRSL.csv", sep = ",")

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

str(BPRSL)

summary(complete.cases(BPRSL))
```

### Graphical look at the data
A graphical look at the data shows at least one value in the treatment 2 group to be the highest throughout the 8 weeks.

```{r, warning=FALSE}
# Check the dimensions of the data
dim(BPRSL)
 
# Plot the BPRSL data
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(name = "Week") + 
  scale_y_continuous(name = "BPRS") 

```

### Fitting Linear Mixed Models to the BPRS data
Looking at the Pr(>|t|) values below it seems that while the bprs score decreases as the weeks progress, the treatment group has no significant effect.

```{r, warning=FALSE}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs ~ week + treatment, BPRSL)
 
# print out a summary of the model
summary(BPRS_reg)

# a scatterplot matrix of the repeated measures of bprs
pairs(BPRSL[,1:4])
```

### Random Intercept and Random Slope Models
This kind of model allows the linear regression fits for each individual to differ in intercept, but also in slope making it possible to account for individual differences in the subjects changing BPRS and also the effect of time. 

The models below are compared using Anova. Neither one shows a significant impact of treatment on bprs.


```{r}
library(lme4)
 
# create a random intercept and random slope models
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
 
# print a summary of the models
summary(BPRS_ref)
summary(BPRS_ref1)


# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

# draw the plot of BPRSL with the bprs values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(name = "Week") + 
  scale_y_continuous(name = "BPRS") 

```


Here we do a random intercept and slope model that allows for a treatment × week interaction.
```{r, warning=FALSE}
# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
 
# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRSL with the bprs values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(name = "Week") + 
  scale_y_continuous(name = "BPRS") 

```

```{r, warning=FALSE}
# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)
 
# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>%
  mutate(Fitted)
 
# draw the plot of BPRSL with the Fitted values of bprs
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(name = "Week") + 
  scale_y_continuous(name = "BPRS") 

```



