# Chapter 2: Regression and model validation

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

### Analysis: Section 1
I started by checking my working directory
```{r}
getwd()
```

Read in the data which includes headers and uses comma as a separator
```{r}
students2014 <- read.table("C:/Users/OMISTAJA/Documents/IODS-project/data/learning2014.txt", header=TRUE, sep=",")
```

Explored the structure of the data to show that it has 166 obs. of  7 variables
```{r}
str(students2014)
```

Explored the structure of the data using dim(), showing that it has 166 rows and 7 columns
```{r}
dim(students2014)
```

**A Brief Description of the Dataset:** 
This data set is from the international survey of Approaches to Learning.
The central measure in this study were questions related to Deep, Surface and Strategic learning answered on a Likert scale. There were 8-12 questions for each category. Combination variables were created from these which were then scaled to the original scales by taking the mean.
Data is also included for Age, attitute toward statistics, exam points, and gender.
Excluded data: observations with exam points of zero were excluded. 


### Analysis: Section 2
A graphical overview of the data using ggplot2 and GGally to make an advanced plot matrix

```{r}
library(ggplot2)
library(GGally)
ggpairs(students2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
```

A summary of the variables in the data
```{r}
summary(students2014)
```

An interpretation of the outputs:
* The majority of the students were female. 
* The majority of the students were young.
* Most of the variables are randomly distributed.
* There was strong correlation between attitude and points for both genders.

```{r}
gender_comparison <- ggplot(students2014, aes(x = attitude, y = points, col = gender, labs(title="Student's attitude versus exam points")))
gender_comparison + ggtitle("Student's attitude versus exam points") + geom_smooth(method = "lm")
```

### Analysis: Section 3 & 4
Three explanatory variables (attitude, strategic learning, surface learning) fitted to a regression model with exam points as the target (dependent) variable:

```{r}
summary(lm(points ~ attitude + stra + surf, data = students2014))
```

Interpretation of the results: Attitude is the only statistically significant predictor of exam points with a t-test p value, Pr(>|t|), less than 0.001. The better people's attitude towards learning was the better they tended to do in the exam.

The same model with the statistically insignificant explanatory variables removed:
```{r}
final_model <- lm(points ~ attitude, data = students2014)
summary("final_model")
```

The Multiple R-squared of the model, 0.2074 (or 20%), shows how much the variance in the points is explained by the attitude. It shows that a good attitude is valuable but that it's far from being the only factor in play.


### Analysis: Section 5
Diagnostic plots: Residuals vs Fitted values, Normal QQ-plot and Residuals vs Leverage. 
```{r}
par(mfrow = c(1,3))
plot(final_model, which = c(1,2,5))
```

The assumptions of a linear regression model are: 
* Errors have constant variance σ^2
* Errors normally distributed 
* Errors not correlated
* The size of a given error does not depend on the explanatory variables 

The variance of the errors are reasonable constant, this is shown by the values in the Residuals vs Fitted values plot being randomly scattered.
The errors are normally distributed, this is shown by most of the values staying along the regression line in the QQ plot.
No one value should have a leverage and while the Residuals vs Leverage plot shows outliers with higher leverage the x-axis is very short so the graphic makes it seem like a bigger problem than it is.

