# Chapter 2: Regression and model validation

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods.
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using.

### Analysis: Section 1
I started by checking my working directory
```{r}
getwd()
```

Read in the data which includes headers and uses comma as a separator
```{r}
students2014 <- read.table("C:/Users/OMISTAJA/Documents/IODS-project/data/learning2014.txt", header=TRUE, sep=",")
```

Explored the structure of the data to show that it has 166 obs. of  7 variables
```{r}
str(students2014)
```

Explored the structure of the data using dim(), showing that it has 166 rows and 7 columns
```{r}
dim(students2014)
```

**A Brief Description of the Dataset:** 
This data set is from the international survey of Approaches to Learning.
The central measure in this study were questions related to Deep, Surface and Strategic learning answered on a Likert scale. There were 8-12 questions for each category. Combination variables were created from these which were then scaled to the original scales by taking the mean.
Data is also included for Age, attitute toward statistics, exam points, and gender.
Excluded data: observations with exam points of zero were excluded. 


### Analysis: Section 2
A graphical overview of the data using ggplot2 and GGally to make an advanced plot matrix

```{r}
library(ggplot2)
library(GGally)
ggpairs(students2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
```

A summary of the variables in the data
```{r}
summary(students2014)
```

An interpretation of the outputs: 
- Describe and interpret the outputs, 
- commenting on the distributions of the variables 
- and the relationships between them.


```{r}
gender_comparison <- ggplot(students2014, aes(x = attitude, y = points, col = gender, labs(title="Student's attitude versus exam points")))
gender_comparison + ggtitle("Student's attitude versus exam points") + geom_smooth(method = "lm")
```


### Analysis: Section 3

### Analysis: Section 4

### Analysis: Section 5
