# Chapter 6: Analysis of longitudinal data

Note: As you read the data sets in your chapter6.Rmd you may have to factor the categorical variables again, as R may not recognise them automatically as factors.

Note that you must SWAP the data sets! :) It is NOT a simple copy & paste of my book!

## 1. Analyses of Chapter 8 of MABS using the RATS data
### Graphs or analysis results 
### Interpretations

```{r}
RATSL <- read.csv("data/RATSL.csv", sep = ",")

names(RATSL)

# Look at the structure of rats
str(RATSL)

# Print out summaries of the variables
summary(RATSL)
```

```{r}
library(dplyr, tidyr, ggplot2)

# Draw the plot. Inspect the plot. See how both the BPRS-score and the variability between individuals decrease over the eight weeks time
ggplot(RATSL, aes(x = week, y = rats, linetype = Group)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$rats), max(RATSL$rats)))

```

## 2. Analyses of Chapter 9 of MABS using the BPRS data. 
### Graphs or analysis results
### Interpretations

```{r}
bprs <- read.csv("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep = " ")

names(bprs)
```


```{r, fig.width=10, fig.height=10, warning=FALSE}
```
